---
title: "Sex disparity of Mutational Signatures"
output: 
  rmdformats::readthedown:
    highlight: kate
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


### Introduction:

1. Somatic mutations found in cancer genomes were originated from different mutational processes
2. Mutation burden varied among and within cancer types
3. The difference in mutation burden imply the deviation of mutation rate
4. Age-related signatures will be focus on given the limitation of exposure time data

### Data preparation：

* [Signature contribution data(FULL_SET.031918)](https://dcc.icgc.org/releases/PCAWG/mutational_signatures/Signatures_in_Samples/SA_Signatures_in_Samples/SA_Full_PCAWG_Attributions)
* [Sample clinical and histology data(August2016_v9)](https://dcc.icgc.org/releases/PCAWG/clinical_and_histology)
* [Patient clinical and histology data(August2016_v9)](https://dcc.icgc.org/releases/PCAWG/clinical_and_histology)

```{r }

library(tidyverse)

# setwd("/Users/jianlongliao/xiaolonggeのhome/Project_Exploration")
setwd("G:/xiaolonggeのhome/ongoing_research/Mutational_Signatures_Research/Exploration")

donor_data <- read_tsv("Data/pcawg_donor_clinical_August2016_v9.txt") # Patient clinical
sam_data <- read_tsv("Data/pcawg_specimen_histology_August2016_v9.txt") # Sample clinical
sig_aty <- read_tsv("Data/SignatureAnalyzer_SNV.activity.FULL_SET.031918.txt") # Signature contribution

```

1. Extract sample neccessary info（<u>cancer type, sex and age at diagnosis</u>）

```{r}

sam_comb_data <- sam_data %>%
  distinct(icgc_specimen_id, .keep_all = TRUE) %>%
  left_join(donor_data, by = "icgc_donor_id") %>%
  filter(!is.na(histology_abbreviation)) %>% 
  select(icgc_specimen_id, histology_abbreviation, donor_sex, donor_age_at_diagnosis)

```

```{r echo=FALSE}
DT::datatable(head(sam_comb_data), options = list(dom = 't'))
```


2. Extract sample signature exposure info

```{r}
sig_aty_df <- sig_aty %>%
  pivot_longer(cols = -X1, names_to = "sample", values_to = "exposure") %>%
  rename(signature = X1) %>%
  separate(sample, into = c("histology_abbreviation", "icgc_specimen_id"), "__") %>%
  select(-histology_abbreviation)
```

```{r echo=FALSE}
DT::datatable(head(sig_aty_df), options = list(dom = 't'))
```

3. Merge sample and signature exposure data

```{r}

sig_comb_data <- sig_aty_df %>%
  left_join(sam_comb_data, by = "icgc_specimen_id")

```

```{r echo=FALSE}
DT::datatable(head(sig_comb_data), options = list(dom = 't'))
```

### Identify age-related signatures in all samples

```{r}

sig_comb_stat <- sig_comb_data %>% 
  group_by(signature) %>% 
  summarise(zero_exp_frac = signif(mean(exposure == 0), 4), 
            min_exp = signif(min(exposure), 4), 
            max_exp = signif(max(exposure), 4),
            mean_exp = signif(mean(exposure), 4),
            median_exp = signif(median(exposure), 4),
            sd_exp = signif(sd(exposure), 4))

```

```{r echo=FALSE}
DT::datatable(sig_comb_stat, options = list(scrollX = TRUE))
```

```{r}

total_rlm_model <- sig_comb_data %>% 
  mutate(signature = fct_inorder(signature)) %>% 
  split(.$signature) %>% 
  map(possibly(~MASS::rlm(exposure ~ donor_age_at_diagnosis, data = .), NA))

```



```{r}

total_lm_p <- sig_comb_data %>% 
  mutate(signature = fct_inorder(signature)) %>% 
  split(.$signature) %>% 
  map_chr(possibly(~summary(lm(exposure ~ donor_age_at_diagnosis, data = .))$coefficients[2, 4], NA))

```


```{r}
total_rlm_p <- total_rlm_model %>% 
  map_chr(possibly(~sfsmisc::f.robftest(., var = 'donor_age_at_diagnosis')$p.value, NA))

```


```{r}

total_rlm_result <- total_rlm_model %>% 
  map(possibly(~summary(.)$coefficients['donor_age_at_diagnosis', c('Value', 'Std. Error')], NA)) %>%
  bind_rows() %>% 
  mutate(signature = unique(sig_comb_data$signature),
         p_value = total_rlm_p, 
         q_value = p.adjust(p_value, method = "BH"),
         q_signi = q_value < 0.05)
 
```





```{r}
total_rlm_result
```




```{r}

# linear regression for signature exposure and age divided by cancer type 

sig_age_rlm <- sig_comb_data %>%
  filter(histology_abbreviation %in% sex_age_stat$histology_abbreviation) %>%
  split(.$histology_abbreviation) %>% # for each cancer type
  map(function(cancer_df){
    cancer_df %>%
      split(cancer_df$donor_sex) %>% # for each sex
      map(function(can_sex_result){
        sig_lm <- can_sex_result %>%
          split(can_sex_result$signature) %>% # for each signature
          map(possibly(~summary(lm(exposure ~ donor_age_at_diagnosis, data = .))$coefficients[2, c(1, 4)], NA)) %>% # linear regression
          bind_rows() %>%
          mutate(signature = unique(can_sex_result$signature),
                 q_value = p.adjust(`Pr(>|t|)`, method = "BH"),
                 q_signi = q_value < 0.05,
                 donor_sex = unique(can_sex_result$donor_sex),
                 histology_abbreviation = unique(can_sex_result$histology_abbreviation))
      }) %>%
      bind_rows()
  }) %>%
  bind_rows()




```




```{r echo=FALSE}
knitr::knit_exit()
```



### Identify sex-biased and age-related signatures







2. Merge sample age and signature exposure data
   * Focus on cancer types which have enough both genders (**> 10 samples**, respectively) to guarantee the statistical power

```{r}

# limited to specific cancer types based on sufficient sex info

sex_stat_lim <- sam_comb_data %>% 
  count(histology_abbreviation, donor_sex) %>% 
  pivot_wider(names_from = donor_sex, values_from = n, values_fill = 0) %>% 
  filter(female >= 10, male >= 10) %>% 
  pivot_longer(-histology_abbreviation, names_to = 'donor_sex', values_to = 'count')


# statistics of age and exposure

exposure_stat <- sig_comb_data %>% 
  filter(!is.na(exposure)) %>%
  group_by(histology_abbreviation, donor_sex) %>%
  summarise(min_exp = min(exposure), 
            max_exp = max(exposure),
            mean_exp = signif(mean(exposure), 4),
            sd_exp = signif(sd(exposure), 4))  

sex_age_stat <- sam_comb_data %>%
  filter(histology_abbreviation %in% sex_stat_lim$histology_abbreviation, !is.na(donor_age_at_diagnosis)) %>% 
  group_by(histology_abbreviation, donor_sex) %>%
  summarise(min_age = min(donor_age_at_diagnosis), 
            max_age = max(donor_age_at_diagnosis),
            mean_age = signif(mean(donor_age_at_diagnosis), 4),
            sd_age = signif(sd(donor_age_at_diagnosis), 4)) %>% 
  left_join(sex_stat_lim, by = c('histology_abbreviation', 'donor_sex')) %>% 
  select(histology_abbreviation, donor_sex, count, everything()) %>% 
  left_join(exposure_stat, by = c('histology_abbreviation', 'donor_sex'))

```

**Table1: Age distribution of satisfactory cancer types**

```{r echo=FALSE}
sex_age_stat %>% DT::datatable(options = list(scrollX = TRUE))
```

3. Conduct simple linear regression analysis

* Simple linear regression between age and signature exposure in a specific cancer type divided by sex

```{r}



# linear regression for signature exposure and age divided by genders and cancer type 

sig_age_lm <- sig_comb_data %>%
  filter(histology_abbreviation %in% sex_age_stat$histology_abbreviation) %>% 
  split(.$histology_abbreviation) %>% # for each cancer type
  map(function(cancer_df){
    cancer_df %>% 
      split(cancer_df$donor_sex) %>% # for each sex
      map(function(can_sex_result){
        sig_lm <- can_sex_result %>%
          split(can_sex_result$signature) %>% # for each signature
          map(possibly(~summary(lm(exposure ~ donor_age_at_diagnosis, data = .))$coefficients[2, c(1, 4)], NA)) %>% # linear regression
          bind_rows() %>%
          mutate(signature = unique(can_sex_result$signature), 
                 q_value = p.adjust(`Pr(>|t|)`, method = "BH"), 
                 q_signi = q_value < 0.05, 
                 donor_sex = unique(can_sex_result$donor_sex), 
                 histology_abbreviation = unique(can_sex_result$histology_abbreviation))
      }) %>%
      bind_rows()
  }) %>% 
  bind_rows()

```

* Search for sex-biased age-related signatures

```{r}

# signatures showing sex disparity in age correlation

sig_age_sexdf <- sig_age_lm %>%
  select(signature, q_signi, donor_sex, histology_abbreviation) %>% 
  pivot_wider(names_from = donor_sex, values_from = q_signi) %>% 
  filter(female != male, !is.na(female), !is.na(male))

sexdf_info <- sig_age_lm %>% 
  semi_join(sig_age_sexdf, by = c('signature', 'histology_abbreviation')) %>%
  mutate(Estimate = signif(Estimate, 4), `Pr(>|t|)` = signif(`Pr(>|t|)`, 4), q_value = signif(q_value, 4)) %>% 
  select(histology_abbreviation, signature, donor_sex, everything()) %>% 
  arrange(histology_abbreviation, signature)

```

**Table2: Sex-biased age-related signatures**

```{r echo=FALSE}
sexdf_info %>% DT::datatable(options = list(scrollX = TRUE))
```



```{r}
# plotting

plot_data <- sig_comb_data %>% 
  semi_join(sig_age_sexdf, by = c('signature', 'histology_abbreviation')) %>% 
  left_join(sexdf_info, by = c('signature', 'histology_abbreviation', 'donor_sex')) %>% 
  mutate(x_coor = 25, y_coor = 4000, `Pr(>|t|)` = str_c('p = ', signif(`Pr(>|t|)`, 3)))

uni_sig <- unique(plot_data$signature)

```


## {.tabset .tabset-pills}

```{r, results='asis'}

for(sig in uni_sig){
  
  # sig <- uni_sig[1]
  sub_data <- plot_data %>% filter(signature %in% sig)
  # up_outlier <- quantile(sub_data$exposure)[4] + 1.5 * IQR(sub_data$exposure)
  # ratio_val <- 100/up_outlier
  
  cat('###',sig,' \n')
  print(ggplot(sub_data, aes(donor_age_at_diagnosis, exposure)) +
       geom_point(col = "darkgreen") +
       geom_smooth(method = 'lm', col = "darkred") +
       coord_fixed(ylim = c(0, 5000), ratio = 1/100) +
       # coord_fixed(ylim = c(0, up_outlier), ratio = ratio_val) +
       geom_text(aes(x_coor, y_coor, label = `Pr(>|t|)`), 
                 data = distinct(sub_data, histology_abbreviation, donor_sex, .keep_all = TRUE), size = 3) +
       facet_grid(donor_sex ~ histology_abbreviation) + 
       theme_bw() +
       theme(axis.text.x = element_blank(), 
             axis.ticks.x = element_blank(), 
             panel.grid = element_blank(), 
             strip.background = element_blank()))
  cat(' \n \n')

}

```


```{r echo=FALSE}
knitr::knit_exit()
```






